\FOO: A framework for efficient source code generation\\ of detection algorithms for the Internet of Things

### 2015-04-17 paper-euspn15.tex
completed text clean-up, squeezed whitespace

\input{packages}

@@

\author[]{Christophe Van Ginneken}
\author[]{Jef Maerien}
\author[]{Christophe Huygens}
\author[]{\\Wouter Joosen}
\author[]{Danny Hughes}

@@

The ability to detect patterns in its usage is what makes \emph{things}
\emph{smart}. The more patterns it can detect, the more accurate its
operational results will be. However, supporting an adequate set of pattern
detection algorithms imposes significant overhead on resource-constrained
devices such as those in the Internet of Things. Manual fusion of the
algorithms reduces resource consumption by optimizing resource sharing, yet,
proves to be time-consuming, repetitive and error-prone. To address this
problem we propose \FOO, a framework for the generation of efficient detection
algorithms. It consists of a domain specific language and a code generator. The
language allows formally describing the intent of detection algorithms and
supports the generator in organizing the source code. As a case study, we apply
\FOO to the generation of an Intrusion Detection System for wireless sensor
nodes. A side-by-side comparison shows that \FOO-generated source code reduces
message passing overhead, execution time and memory footprint in comparison to
sequential calls into stand-alone implementations of the algorithms.

@@

IoT, Domain Specific Languages, Source Code Generation, Source Code
Optimization, Intrusion Detection, Context Awareness

@@

\vspace*{-10pt}

\section{Introduction}

The Internet of Things (IoT), the mother of all ubiquitous and pervasive
networks, populates our daily lives with millions of internet-connected
\emph{smart things}: from smart watches that assist visually impaired
persons\cite{porzi2013smart} to monitoring systems for flood hazardous
areas\cite{hughes2006gridstix}. Their applications cannot be more diverse,
however they share one common concern: each of these \emph{things} is
resource-constrained, relying on a battery and operating with limited
processing power and memory.

Although their applications at first seem unrelated, under the hood they share
a common paradigm: pattern recognition. With the introduction of \emph{smart
things}, we want to augment our lives with information that is hard, or at
least time and resource consuming, to obtain otherwise. The ability to detect
patterns in its usage or its applied context, is what makes these \emph{things}
\emph{smart}. A watch that recognizes gestures can perform certain actions on
behalf of its wearer. That same watch can also detect a wet floor and notify
this hazard.

Monitoring human activities is an active research field for more than a
decade\cite{lara2013survey}, but with the advent of the IoT it now moves into
the real world. Equipped with wearables with numerous sensors, the
possibilities seem endless, but so are the number of detection and recognition
algorithms that have to run simultaneously on these small \emph{things}, to
support our augmented reality.

Another prime example of a class of algorithms that inhibit the same properties
related to pattern recognition is Intrusion Detection (ID). The vast number of
security related opportunities for attackers makes way for an equally sized
amount of ID algorithms, all with a similar structure, but largely independent
operation.

@@

The contributions of this paper are threefold: The first contribution consists
in the formulation of a design pattern for detection/recognition algorithms for
IoT devices. This defines the target algorithms that fit our proposed
framework. The second contribution is a framework called \FOO.\@ It enables the
formal description of algorithms on a functional level and provides a code
generator producing source code for a given platform and configuration. This
way, \FOO also addresses the heterogeneous nature of both IoT and the
algorithms. The third contribution introduces functional code fusion (FCF) as a
source code generation paradigm to address the combination of algorithms. FCF
identifies common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show reduced usage of the wireless
radio, execution time and memory usage.

@@

The remainder of this paper proceeds as follows: section \ref{problems}
identifies problems that arise when combining multiple algorithms in the
context of resource-constrained devices. Section \ref{pattern} formulates a
pattern common to the detection/recognition algorithms in the IoT context.
Section \ref{design} introduces our framework, its design and components. In
section \ref{evaluation} we apply our solution to case study related to ID,
showing an actual implementation and evaluation of the experimental results.
Section \ref{related} explores related work in the field of DSLs and code
generation. Finally, section \ref{conclusion} summarizes our findings, draws
conclusions and identifies topics for future work.

@@

Simply taking existing implementations of algorithms and calling them in
sequence, to allow them to perform their functionality, will in most cases
result in suboptimal code. Three prime examples on the source code level
illustrate this:

\vspace{-2mm}

\begin{description}

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can cause overhead. An example is
  the handling of messages: each algorithm parses each observed message on the
  network, processing the same byte stream, performing comparable operations to
  find similar patterns. These loops can also hold calls to poll sensor for
  their values. This results in an execution overhead, putting a strain on the
  power consumption.

\vspace{-2mm}

  \item[Common data] cross-cuts the different algorithms, that have to store
  the same information over and over again within their own scope. This results
  in a lot of duplicated information in memory, raising the hardware
  requirements.

\vspace{-2mm}

@@

\vspace{-2mm}

\begin{description}

@@

\vspace{-2mm}

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This partially explains why
  there are few implementations of algorithms readily available to developers:
  there is no single platform to target for researchers, which lowers the
  opportunity to reuse prior work.

@@

\section{A Design Pattern for Detection/Recognition Algorithms}
\label{pattern}

The source code level problems that arise when combining multiple algorithms,
as presented in section \ref{problems}, are in fact inherent to the design
pattern that is the foundation for the type of algorithms we identify in
typical detection/recognition applications in IoT applications. The formulation
of such a design pattern will allow us to define a strategy for the
optimization of the combination of these algorithms. We start by extracting two
common activities of these algorithms: the collection of sensor values and the
evaluation of thresholds. Periodically collecting sensor values requires
preparing the hardware for value extraction, waiting for the value to become
available, reading the actual bytes, interpreting them into a meaningful value
and recording them for further analysis. This can be as simple as reading a
single byte, but can also be as complex as parsing a network message.

A single recorded sensor value hardly ever results in a decision. The
combination, often over time, of several of these values and the evaluation of
thresholds will be key to decide if a change in the environment requires an
action. Updating such statistics and evaluating the corresponding thresholds
should be distinct operations. Therefore this is a separate function that is
also periodically executed.

This decoupling of detection and decision is the core of the design pattern for
ID algorithms. Listing \ref{alg:pattern} presents the design pattern in
pseudo-code. Applying the pattern in the context of multiple algorithms,
extends its scope and a pattern such as described in listing
\ref{alg:application} will typically appear. The pattern calls the individual,
implementations in sequence in two situations: first, to collect sensor
readings, and secondly, at a regular interval to allow for internal
housekeeping, such as threshold evaluation.

We clearly see the three aforementioned problems manifest itself: first, the
nesting of loops causes redundant executions of potentially the same sensor
readings inside the individual algorithms' $collect\_data$ functions. Second,
each individual algorithm stores the same sensor values in memory. Third,
scattered calls to e.g. network functions (in case of distributed algorithms)
lead to suboptimal use of the wireless radio.

Implementing the solution from scratch, it would be possible to identify these
problems and reorganize the code in such a way that sensor values are only read
once, stored in memory once and network messages parsed once. In section
\ref{problems} we identified two operational problems that show that this is
not feasible or at least suboptimal. We therefore present \FOO, a framework
that enables formally describing such algorithms and fusing common data and
functions in an automated way, to overcome both these problems.

\vspace{-2mm}

\begin{algorithm}[b]
  \caption{Detection/Recognition Algorithm Pattern}
  \label{alg:pattern}
  \begin{algorithmic}
    \Require{readings: global variables to hold collected sensor data for the given algorithm}
    \Require{devices: global variable to information about neighbor devices} \Comment{optional in case of distributed algorithms}
    \Function{collect\_data}{}
      \ForEach{$sensor \in required\_sensors$}
        \Let{values[]}{$sensor$}   \Comment{collect sensor data}
        \Let{$reading_{sensor}$}{\Call{interpret}{values[]}}  \Comment{update interpreted sensor info}
      \EndFor
      \State \Call{send}{$devices_x$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
    \EndFunction
    \Function{do\_housekeeping}{}
      \ForEach{$reading \in readings$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
        \If{$reading > \dots$} \Comment{validate recorded value}
          \State \dots \Comment{take actions}
          \State \Call{send}{$device_x$, ``info''} \Comment{optionally communicate} \label{alg:id-algo-pattern-send2}
        \EndIf
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[b]
  \caption{Application pattern of multiple algorithms}
  \label{alg:application}
  \begin{algorithmic}
    \ForEach{$algorithm \in algorithms$}
      \State \Call{algorithm.collect\_data}{}
    \EndFor
    \ForEach{$algorithm \in algorithms$}        \Comment{at a given interval}
      \State \Call{algorithm.do\_housekeeping}{}
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to detection
algorithms, but conflict with the target environment, the IoT\@. \FOO provides
a framework to efficiently implement these algorithms for the IoT and offers a
solution for all identified problems. It consists of a 1) DSL, 2) source code
generator and 3) software library. Figure \ref{fig:design} visualizes the
design.

\begin{figure}[b]
  \centering
  \scalebox{0.75}{
\begin{tikzpicture}[
  node distance=9mm,
  ,>=stealth,very thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)     [user]                                      {DSL};
  \node (cg)      [representation,right=of dsl,align=center]  {Code\\Generator};
  \node (code)    [generated,right=of cg]                     {code};
  \node (comp)    [representation,right=of code]              {Compiler};
  \node (image)   [generated,right=of comp]                   {image};

@@

The DSL enables researchers to formally describe algorithms, without focusing
on any specific language or platform. A code generator can avoid the identified
problems concerning loops, common data and network access, as the formal
description provides the underlying intent, not the technical implementation.
It can also target different languages and platforms, while reusing a software
library that provides services including parsing, network access wrapping,\dots
It applies the Functional Code Fusion (FCF) paradigm to identify the hotspots
for optimization. A key component of FCF is the introduction of
\emph{functional domains} that harbor common functionality.

Using a DSL and code generator defines a development strategy. We will first
position \FOO and motivate its design. Second, we will look at the principal
design components of the DSL.\@ Third, we discuss the structure of the code
generator and introduce its software library, that offers a way to deal with
for example the network access issue.

\vspace{-1mm}
\subsection{\FOO as a Development Strategy}
\label{positioning}

The introduction of a code generator in the development process changes the way
of working. The actual source code is no longer the driving force, controlled
by developers. It is literally an intermediate representation between code
generator and compiler. A code generator constructs this representation using
three components, as shown in figure \ref{fig:design}: 1) the description of
the algorithm in the DSL, 2) information about the functional domains, and 3)
information about the platform and target language. At the level of the
compiler, a software library offloads specialized implementations and allow
custom extensions. The applied grayscale indicates the components' platform
dependency. This hierarchy is important from a development effort point of
view. The level of dependency on the target platform dictates the level of
effort to implement changes. This is important in relation to the scalability
of the framework itself.

\vspace{-1mm}
\subsection{\FOO as a Domain Specific Language}
\label{dsl-design}

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \FOO's DSL and model focus on avoiding constructions that
result in problems concerning loops, common data and the use of the network. To
do this, \FOO uses an event-based approach to defining functionality.
Describing algorithms in function of events that occur will remove the
technical descriptions and allows FCF to identify the actual intent of the
algorithms with respect to the identified source code problems.

\vspace{-1mm}
\subsubsection{Loops}

@@

\vspace{-1mm}
\subsubsection{Common Data}

Different detection algorithms typically deal with the same classes of
information, be it sensor data or network messages. Functional domains
consolidate these and store and manage this data in a unified way. \FOO
provides the concept of such domains as plug-ins to the DSL, extending it with
domain specific storage and logic. For example in the case of distributed
algorithms, such as ID algorithms, a \emph{Nodes} domain will manage
information about neighbor nodes, exposing them as global objects to the
algorithms, relieving them from managing these themselves. Functional domains
are accessible as collections, allowing execution of functions in their scope,
eliminating the need for explicit loops.

\vspace{-1mm}
\subsubsection{Unified Messaging}
\label{dsl-unified-msg}

Access to the network, scattered around an algorithm, causes the wireless radio
to be more active than needed. Delaying the sending of messages and grouping
them can lower this overhead. Sending and receiving of network messages is
tightly integrated with the previously mentioned concepts of event handling and
for example the Nodes domain. An underlying software library can handle all
interaction with the network, triggering functions by means of events when
parsing new messages and exposing an API to send messages through the Nodes
domain. This approach gives the framework complete control over communication
and allows for optimized marshaling and grouping of messages.

\vspace{-1mm}
\subsection{\FOO as a Code Generator}

To convert the DSL-based ID algorithms into production-ready code, \FOO applies
code generation techniques. Although different approaches are applicable, like
template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the required transformations demand a
richer model. A semantic model \cite{fowler2010domain}, tied closely to the DSL
itself, can fulfill this role. Semantic transformations evolve the model into a
more syntax-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations can further evolve the model into source code. Figure
\ref{fig:code-generation} shows the process that \FOO implements.

\begin{figure}[b]
  \centering
  \scalebox{.75}{
\begin{tikzpicture}[
  node distance=7mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)  [resource]                                 {DSL};
  \node (sm)   [representation,right=of dsl,align=center] {Semantic\\Model};
  \node (cm)   [representation,right=of sm,align=center]  {Syntactic\\Model};
  \node (code) [resource,right=of cm]        {Code};

  \draw [->] (dsl) -> (sm)   node [midway,above=15pt]              {parse};
  \draw [->] (sm)  -> (cm)   node [midway,below=10pt,align=center] {semantic\\transform}
                             node [midway,above=15pt,align=center] {FCF};
  \draw [->] (cm)  -> (code) node [midway,above=15pt]              {emit code};

@@

\vspace{-1mm}
\subsubsection{Semantic Model}

@@

\begin{figure}[b]
  \centering
  \scalebox{.75}{
\begin{tikzpicture}[
  edge from parent fork down,
  sibling distance=5mm, level distance=20mm, node distance=6mm
  ]
  \node (model) [class,fill=black!10!white] {Model}
      child {node (module) [class,fill=black!10!white] {Module} edge from parent [composite]
        child {node (strategy) [class,align=center,xshift=-10.5mm,fill=black!10!white] {Execution\\Strategy} edge from parent [composite]
          child {node(every) [class,xshift=-9mm,fill=black!10!white] {Interval} edge from parent [subclass]}
          child {node (when) [class,xshift=9mm,fill=black!10!white] {Event} edge from parent [subclass]}
        }
        child {node (function) [class,xshift=10.5mm] {Function} edge from parent [composite]}
      };
  \node (constant)  [class,right=of module]                {Constant};
  \node (scope)     [class,left=of strategy]               {Scope};
  \node (statement) [class,right=of function]              {Statement};
  \node (domain)    [class,left=of module,xshift=-13mm]    {Functional\\Domain};

@@

  \draw [composite]  (domain)   -- (scope);
  \draw [dependency] (strategy) -- (function);
\end{tikzpicture}
}
\caption{Partial Semantic Model for \FOO.\@ Shaded classes indicate the
\emph{event/scheduling} construction around functions to allow interpretation
of the intended functionality.}
\label{fig:meta-model}
\end{figure}

\vspace{-1mm}
\subsubsection{Functional Code Fusion (FCF)}

@@

\vspace{-1mm}
\subsubsection{Syntactic Model}

The syntactic model is like an abstract syntax tree (AST) for a virtual,
expressive language. It contains constructs and ideas found in different
languages and programming paradigms. No single existing programming language
supports all features as is. Syntactic transformations therefore translate the
unsupported constructs into comparable constructs supported by the targeted
platform and language. Depending on the target platform, some constructions
aren't transformed into actually corresponding code. In selected cases, the
generator uses a software library. For example when it needs to parse incoming
messages or schedule tasks. During the generation process, it looks for
patterns that indicate any of these actions and transforms them into calls to
the \FOO software library.

\vspace{-1mm}
\subsection{\FOO as a Software Library}
\label{software-lib-design}

A software library allows offloading of common implementation aspects, thus
avoiding small discrepancies in code and suboptimal use of shared resources.
Previously we illustrated this approach with a case for unified messaging:
parsing of incoming messages and grouping of outgoing messages. More common
functionality isn't generated verbatim either. The execution of functions based
on a schedule or the implementation of higher level conceptual functions are
calls into the standard software library. Optimization of such algorithms is
not within the scope of \FOO.\@ This gives freedom of choice to the users of
the framework to select appropriate implementations for this standard
functionality.

\section{Implementation and Evaluation: Case Study on Intrusion Detection in Wireless Sensor Networks}
\label{evaluation}

In wired networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets. Still, attacks on services can pass unnoticed. This
is where an Intrusion Detection System (IDS) comes into play: monitoring all
traffic that passes through the firewall, looking for patterns of malicious
activity\cite{denning1987intrusion}. In meshed networks it is not possible for
a single point in the network to oversee all traffic
\cite{mishra2004intrusion}. Every device has to implement its own IDS.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the physical layer up to the
application layer. Each layer presents different opportunities for launching an
attack. For each of these attacks an algorithm needs to look for patterns or
anomalies. An IDS therefore requires a many algorithms to achieve adequate
coverage.

\vspace{-1mm}
\subsection{Implementation}

@@

We started from a basic application that measures light intensity. The
application was an includable piece of code, allowing integration in a both a
manual implementation and generated code base. On top of this baseline we added
two intrusion detection related algorithms: a watchdog
\cite{mishra2004intrusion}, that allows nodes to validate each other's
presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
checks if a parent node is cooperative and actually forwards messages.

\vspace{-1mm}
\subsection{Analytical Results}

@@

\vspace{-1mm}
\subsection{Experimental Results}

>>>>

\input{packages}
\input{squeeze}

@@

\author{Christophe Van Ginneken}
\author{Jef Maerien}
\author{Christophe Huygens}
\author{Wouter Joosen}
\author{Danny Hughes}

@@

The ability to detect patterns in its usage is what makes \emph{things}
\emph{smart}. The more patterns it can detect, the more accurate its
operational results will be. However, supporting an adequate set of pattern
detection algorithms imposes significant overhead on resource-constrained
devices such as those in the Internet of Things. Manual fusion of the
algorithms reduces resource consumption by optimizing resource sharing, yet,
proves to be time-consuming, repetitive and error-prone. To address this
problem we propose \FOO, a framework for the generation of efficient
combinations of detection algorithms. It consists of a domain specific language
and a code generator. The language allows formally describing the intent of
detection algorithms and supports the generator in organizing the source code.
As a case study, we apply \FOO to the generation of an intrusion detection
system for wireless sensor nodes. A side-by-side comparison shows that
\FOO-generated source code reduces message passing overhead, execution time and
memory footprint in comparison to sequential calls into standalone
implementations of the algorithms.

@@

Internet of Things, Domain Specific Languages, Source Code Generation, Source
Code Optimization, Intrusion Detection, Context Awareness

@@

\section{Introduction}
\label{introduction}

The Internet of Things (IoT), the mother of all ubiquitous and pervasive
networks, populates our daily lives with millions of internet-connected
\emph{smart things}: from smart watches that assist visually impaired
persons\cite{porzi2013smart} to monitoring systems for waste management in our
\emph{smart cities}\cite{zanella2014internet}. Their applications cannot be
more diverse, however they share one common concern: each of these
\emph{things} is resource-constrained, relying on a battery and operating with
limited processing power and memory.

Although their applications at first seem unrelated, under the hood they share
a common paradigm: pattern recognition. With the introduction of \emph{smart
things}, we want to augment our lives with information that is hard, or at
least time and resource consuming, to obtain otherwise. The ability to detect
patterns in its usage or its applied context, is what makes these \emph{things}
\emph{smart}. A watch that recognizes gestures can perform certain actions on
behalf of its wearer. That same watch can also detect a wet floor and notify
this hazard\cite{porzi2013smart}.

Monitoring human activities is an active research field for more than a
decade\cite{lara2013survey}, but with the advent of the IoT it now moves into
the real world at an increased pace. Equipped with wearables, hosting numerous
sensors, the possibilities seem endless, but so are the number of detection and
recognition algorithms that have to run simultaneously on these small
\emph{things}, to support our augmented reality.

Another prime example of a class of algorithms that inhibit the same properties
is Intrusion Detection (ID). The large number of security related opportunities
for attackers requires an equally sized amount of ID algorithms, all with a
similar structure, still independent in their operation.

@@

The contributions of this work are threefold: the first contribution consists
in the formulation of a design pattern for detection/recognition algorithms for
IoT devices. This defines the target algorithms that fit our proposed
framework. The second contribution is a framework called \FOO.\@ It enables the
formal description of algorithms on a functional level and provides a code
generator producing source code for a given platform and configuration. This
way, \FOO also addresses the heterogeneous nature of both IoT and the
algorithms. The third contribution introduces functional code fusion (FCF) as a
source code generation paradigm to address the combination of algorithms. FCF
identifies common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show reduced usage of the wireless
radio, execution time and memory usage.

@@

The remainder of this paper proceeds as follows: section \ref{problems}
identifies problems that arise when combining multiple algorithms in the
context of resource-constrained devices. Section \ref{pattern} formulates a
pattern common to detection/recognition algorithms in the IoT context. Section
\ref{design} introduces our framework, its design and components. In section
\ref{evaluation} we apply it to a case study related to ID and evaluate the
results. Section \ref{related} explores related work in the field of DSLs and
code generation. Finally, section \ref{conclusion} summarizes our findings,
draws conclusions and identifies topics for future work.

@@

Simply taking existing implementations of algorithms and calling them in
sequence, will in most cases result in suboptimal code. Three prime examples on
the source code level illustrate this:

\begin{description}[style=unboxed,leftmargin=0cm]

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can cause overhead. An example is
  the handling of messages: each algorithm parses each observed message on the
  network, processing the same byte stream, performing comparable operations to
  find similar patterns. Polling sensors for their value is another example.
  This redundancy results in an execution overhead, putting a strain on the
  power consumption.

  \item[Common data] cross-cuts the different algorithms, that have to store
  the same information over and over again within their own scope. Typical
  examples here include information about neighboring nodes, sensor readings,
  etc.\ This results in a lot of duplicated information in memory, raising the
  hardware requirements.

@@

\begin{description}[style=unboxed,leftmargin=0cm]

@@

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This partially explains why
  there are hardly any implementations of algorithms readily available to
  developers: there is no single platform to target for researchers, which
  lowers the opportunity to reuse prior work.

@@

\section{A design pattern for detection/recognition algorithms}
\label{pattern}

The source code level problems that arise when combining multiple algorithms,
as presented in section \ref{problems}, are in fact inherent to the design
pattern that forms the foundation for the type of algorithms we identify in
typical detection/recognition applications in an IoT context. The formulation
of such a design pattern will allow us to define a strategy for the
optimization of the combination of these algorithms. We start by extracting two
common activities: the collection of sensor values and the evaluation of
thresholds. Periodically collecting sensor values requires preparing the
hardware for value extraction, waiting for the value to become available,
reading the bytes from the sensor, interpreting these bytes and recording this
interpretation for further analysis. This can be as simple as reading a single
byte, but can also be as complex as parsing a network message.

A single sensor value hardly ever results in a decision. The combination, often
over time, of these values and the evaluation of thresholds will be key to
decide if a change in the environment requires an action. Updating such
statistics and evaluating the corresponding thresholds should be distinct
operations. This decoupling of detection and decision is the core of the design
pattern for detection algorithms. Listing \ref{alg:pattern} presents the design
pattern in pseudo-code. Applying the pattern in the context of multiple
algorithms, extends its scope and an application pattern such as described in
listing \ref{alg:application} appears.

We see the three aforementioned problems manifest themselves: first, nesting of
loops causes redundant executions of potentially the same sensor readings
inside individual algorithms' $collect\_data$ functions. Second, each
individual algorithm stores the same sensor values in memory. Third, scattered
calls to network functions (in case of distributed algorithms) lead to
suboptimal use of the wireless radio.

When implementing the solution from scratch, it would be possible to identify
these problems and reorganize the code in such a way that it reads and stores
sensor values in memory once or centrally parses network messages. In section
\ref{problems} we identified two operational problems that show that this is
not feasible, or at least suboptimal. We therefore present \FOO, a framework
that enables formally describing such algorithms and fusing common data and
functions in an automated way, to overcome both these problems.

\begin{algorithm}[t]
  \small
  \caption{Detection/Recognition Algorithm Pattern}
  \label{alg:pattern}
  \begin{algorithmic}
    \Require{readings: global variables to hold collected sensor data for the given algorithm}
    \Require{devices: global variable to information about neighbor devices} \Comment{optional in case of distributed algorithms}
    \Function{collect\_data}{}
      \ForEach{$sensor \in required\_sensors$}
        \Let{values[]}{$sensor$}   \Comment{collect sensor data}
        \Let{$reading_{sensor}$}{\Call{interpret}{values[]}}  \Comment{update interpreted sensor info}
      \EndFor
      \State \Call{send}{$devices_x$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
    \EndFunction
    \Function{do\_housekeeping}{}
      \ForEach{$reading \in readings$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
        \If{$reading > \dots$} \Comment{validate recorded value}
          \State \dots \Comment{take actions}
          \State \Call{send}{$device_x$, ``info''} \Comment{optionally communicate} \label{alg:id-algo-pattern-send2}
        \EndIf
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[t]
  \small
  \caption{Application pattern of multiple algorithms}
  \label{alg:application}
  \begin{algorithmic}
    \ForEach{$algorithm \in algorithms$}
      \State \Call{algorithm.collect\_data}{}
    \EndFor
    \ForEach{$algorithm \in algorithms$}        \Comment{at a given interval}
      \State \Call{algorithm.do\_housekeeping}{}
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

The previous sections identified loops, common data and the network as three
important code level problems. These problems are inherent to detection
algorithms, but conflict with the target environment, the IoT\@. \FOO provides
a framework to efficiently implement these algorithms and offers a solution for
all identified problems. It consists of a 1) DSL, 2) source code generator and
3) software library.

The DSL enables researchers to formally describe algorithms, without focusing
on a specific language or platform. The code generator can avoid the identified
problems concerning loops, common data and network access, because the formal
description provides the underlying intent, not the technical implementation.
It applies Functional Code Fusion (FCF) to identify hotspots for optimization.
A key component of FCF are \emph{functional domains} that harbor common
functionality.

We will first position \FOO as a development strategy and motivate its design.
Second, we will look at the principal design components of the DSL.\@ Third, we
discuss the structure of the code generator and introduce its software library.

\subsection{\FOO as a development strategy}
\label{positioning}

The introduction of a code generator in the development process changes the way
of working: the source code is no longer the driving force, controlled by
developers. It is literally an intermediate representation between code
generator and compiler. A code generator constructs this representation using
three components, as shown in figure \ref{fig:design}: 1) the description of
the algorithm in the DSL, 2) information about the functional domains, and 3)
information about the platform and target language. At the level of the
compiler, a software library offloads specialized implementations and allows
custom extensions. The applied grayscale indicates the components' platform
dependency. This hierarchy is important from a development effort point of
view. The level of dependency on the target platform dictates the level of
effort to implement changes. This is important in relation to the scalability
of the framework itself.

\begin{figure}[t]
  \centering
  \scalebox{0.70}{
\begin{tikzpicture}[
  node distance=9mm,
  ,>=stealth,very thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)     [user]                                      {DSL};
  \node (cg)      [representation,right=of dsl,align=center]  {Code\\Generator};
  \node (code)    [generated,right=of cg]                     {code};
  \node (comp)    [representation,right=of code]              {Compiler};
  \node (image)   [generated,right=of comp]                   {image};

@@

\subsection{\FOO as a domain specific language}
\label{dsl-design}

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \FOO's DSL and model focus on avoiding constructions that
result in problems concerning loops, common data and the use of the network. To
do this, \FOO uses an event-based approach to defining functionality.
Describing algorithms based on events removes the technical interpretations and
allows FCF to identify the intent of algorithms and avoid the identified source
code problems.

\subsubsection{Loops}

@@

\subsubsection{Common data}

Different detection algorithms typically deal with the same classes of
information, be it sensor data or network messages. Functional domains
consolidate these and store and manage this data in a unified way. \FOO
provides the concept of such domains as plug-ins to the DSL, extending it with
domain specific storage and logic. For example in the case of distributed
algorithms a \emph{Nodes} domain will manage information about neighbor nodes,
exposing them as global objects to the algorithms, relieving them from managing
these themselves. Functional domains are also accessible as collections,
allowing execution of functions in their scope, again eliminating the need for
explicit loops.

\subsubsection{Unified messaging}
\label{dsl-unified-msg}

Access to the network, scattered around an algorithm, causes the wireless radio
to be more active than needed. Delaying the sending of messages and grouping
them can lower this overhead. Sending and receiving of network messages is
tightly integrated with the previously mentioned concepts of event handling and
for example the \emph{Nodes} domain. An underlying software library can handle
all interaction with the network, triggering functions by means of events when
parsing messages and exposing an API to send messages. This approach gives the
framework complete control over communication and allows for optimized
marshaling and grouping of messages.

\subsection{\FOO as a code generator}

To convert the DSL-based ID algorithms into production-ready code, \FOO applies
code generation techniques. Although different approaches are applicable, like
template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the required transformations demand a
richer model. A semantic model \cite{fowler2010domain}, tied closely to the DSL
itself, can fulfill this role. Semantic transformations evolve the model into a
more syntax-oriented model, comparable to the idea of CodeDom. Syntactic
transformations can further evolve the model into source code. Figure
\ref{fig:code-generation} shows this process.

\begin{figure}[t]
  \centering
  \scalebox{.70}{
\begin{tikzpicture}[
  node distance=7mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)  [resource]                                 {DSL};
  \node (sm)   [representation,right=of dsl,align=center] {Semantic\\Model};
  \node (cm)   [representation,right=of sm,align=center]  {Syntactic\\Model};
  \node (code) [resource,right=of cm]                     {Source Code};

  \draw [->] (dsl) -> (sm)   node [midway,above=15pt]              {parse};
  \draw [->] (sm)  -> (cm)   node [midway,below=10pt,align=center] {semantic\\transform}
                             node [midway,above=15pt,align=center] {FCF};
  \draw [->] (cm)  -> (code) node [midway,above=15pt]              {emit};

@@

\subsubsection{Semantic model}

@@

\begin{figure}[b]
  \centering
  \scalebox{.70}{
\begin{tikzpicture}[
  edge from parent fork down,
  sibling distance=5mm, level distance=20mm, node distance=6mm
  ]
  \node (model) [class,fill=black!10!white] {Model}
      child {node (module) [class,fill=black!10!white] {Module} edge from parent [composite]
        child {node (strategy) [class,align=center,xshift=-10.5mm,fill=black!10!white] {Execution\\Strategy} edge from parent [composite]
          child {node(every) [class,xshift=-9mm,fill=black!10!white] {Interval} edge from parent [subclass]}
          child {node (when) [class,xshift=9mm,fill=black!10!white] {Event} edge from parent [subclass]}
        }
        child {node (function) [class,xshift=10.5mm] {Function} edge from parent [composite]}
      };
  \node (constant)  [class,right=of module]                {Constant};
  \node (scope)     [class,left=of strategy]               {Scope};
  \node (statement) [class,right=of function]              {Statement};
  \node (domain)    [class,left=of module,xshift=-13mm]    {Functional\\Domain};

@@

  \draw [composite]  (domain)   -- (scope);
  \draw [dependency] (strategy) -- (function);
\end{tikzpicture}
}
\caption{Partial Semantic Model for \FOO.\@ Shaded classes indicate the
\emph{event/scheduling} construction around functions.}
\label{fig:meta-model}
\end{figure}

\subsubsection{Functional code fusion (FCF)}

@@

\subsubsection{Syntactic model}

The syntactic model is like an abstract syntax tree for a virtual, expressive
language. It contains constructs and ideas found in different languages and
programming paradigms. No single existing programming language supports all
features as is. Syntactic transformations therefore translate the unsupported
constructs into comparable constructs supported by the targeted platform and
language. Depending on the target platform, some constructions aren't
transformed into actually corresponding code. In selected cases, the generator
uses a software library. For example when it needs to parse incoming messages
or schedule tasks. During the generation process, it looks for patterns that
indicate any of these actions and transforms them into calls to the \FOO
software library.

\subsection{\FOO as a software library}
\label{software-lib-design}

A software library allows offloading of common implementation aspects, thus
avoiding small discrepancies in code and suboptimal use of shared resources.
Previously we illustrated this approach with a case for unified messaging:
parsing of incoming messages and grouping of outgoing messages. More common
functionality isn't generated verbatim either. A standard software library
provides functionality for the execution of functions based on a schedule or
the implementation of higher level conceptual functions. Optimization of these
type of algorithms is not within the scope of \FOO.\@ This further gives
freedom of choice to the users of the framework to select appropriate
implementations for this standard functionality.

\section{
Implementation and evaluation:
a case study on intrusion detection in wireless sensor networks
}
\label{evaluation}

In wired networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets. Still, attacks on services can pass unnoticed. This
is where an Intrusion Detection System (IDS) comes into play: monitoring all
traffic that passes through the firewall, looking for patterns of malicious
activity\cite{denning1987intrusion}. In meshed networks it is not possible for
a single point in the network to oversee all traffic
\cite{mishra2004intrusion}. So every device has to implement its own IDS.\@

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the physical layer up to the
application layer. Each layer presents different opportunities for launching an
attack. For each of these attacks an algorithm needs to look for patterns or
anomalies. An IDS therefore requires a lot of algorithms to achieve adequate
coverage.

\subsection{Implementation}

@@

We started from a basic application that measures light intensity. The
application was an includable piece of code, allowing integration in both a
manual implementation and generated code base. On top of this baseline we added
two intrusion detection related algorithms: a watchdog
\cite{mishra2004intrusion}, that allows nodes to validate each other's
presence, and a reputation-based algorithm \cite{ganeriwal2008reputation}, that
checks if a parent node is cooperative and actually forwards messages.

\subsection{Analytical results}

@@

\subsection{Experimental results}
